Type-Safe Modular Parsing

### 2017-04-24 SLE_17/sections/CaseStudy.tex
Revise code (up to section 4) to fit the narrower column

\begin{figure}[ht]
\lstinputlisting[linerange=15-33]{code/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_VARAPP
\caption{The \inlinecode{VarApp} component in case study.}\label{fig:casestudy-varapp}
\end{figure}

@@

\lstinputlisting[linerange=9-11]{code/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_EPARSER

@@

\begin{figure}[t]
\lstinputlisting[linerange=37-64]{code/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_UNTYPED
\caption{Build the \inlinecode{Untyped} calculus by composing to language components.}\label{fig:casestudy-untyped}
\end{figure}

@@

\lstinputlisting[linerange=68-78]{code/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_DEMO

>>>>

\begin{figure}[ht]
\lstinputlisting[linerange=15-33]{code/src/papercode/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_VARAPP
\caption{The \inlinecode{VarApp} component in case study.}\label{fig:casestudy-varapp}
\end{figure}

@@

\lstinputlisting[linerange=9-11]{code/src/papercode/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_EPARSER

@@

\begin{figure}[t]
\lstinputlisting[linerange=37-64]{code/src/papercode/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_UNTYPED
\caption{Build the \inlinecode{Untyped} calculus by composing to language components.}\label{fig:casestudy-untyped}
\end{figure}

@@

\lstinputlisting[linerange=68-78]{code/src/papercode/Sec6CaseStudy/Code1.scala}% APPLY:linerange=CASESTUDY_DEMO

### 2017-04-24 SLE_17/sections/Discussion.tex
Revise code (up to section 4) to fit the narrower column

\begin{figure}[ht]
\lstinputlisting[linerange=31-59]{code/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
\caption{Code for parsing a language with multi-sorted syntax for types and expressions.}
\label{fig:multi}
\end{figure}

@@

\lstinputlisting[linerange=63-66]{code/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX_CLIENT
The code works as expected to parse and pretty-print the result.

@@

\lstinputlisting[linerange=76-81]{code/Sec5/Code1.scala}% APPLY:linerange=BASEPARSER_UNTYPEDLAM

@@

\lstinputlisting[linerange=85-90]{code/Sec5/Code1.scala}% APPLY:linerange=EXTPARSER_TYPEDLAM

@@

\begin{figure}[t]
\lstinputlisting[linerange=113-132]{code/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VAREXPR
\lstinputlisting[linerange=136-156]{code/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_TYPEDLAM
\caption{Language components \lstinline{VarExpr} (\huang{todo}left) and \lstinline{TypedLam} (right).}\label{fig:lng-components}
\end{figure}

@@

\begin{figure}[t]
\lstinputlisting[linerange=160-170]{code/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR
\caption{Reuse two language components \lstinline{VarExpr} and \lstinline{TypedLam}.}\label{fig:compose-components}
\end{figure}

>>>>

\begin{figure}[ht]
\lstinputlisting[linerange=31-59]{code/src/papercode/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
\caption{Code for parsing a language with multi-sorted syntax for types and expressions.}
\label{fig:multi}
\end{figure}

@@

\lstinputlisting[linerange=63-66]{code/src/papercode/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX_CLIENT
The code works as expected to parse and pretty-print the result.

@@

\lstinputlisting[linerange=76-81]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=BASEPARSER_UNTYPEDLAM

@@

\lstinputlisting[linerange=85-90]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=EXTPARSER_TYPEDLAM

@@

\begin{figure}[t]
\lstinputlisting[linerange=113-132]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VAREXPR
\lstinputlisting[linerange=136-156]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_TYPEDLAM
\caption{Language components \lstinline{VarExpr} (\huang{todo}left) and \lstinline{TypedLam} (right).}\label{fig:lng-components}
\end{figure}

@@

\begin{figure}[t]
\lstinputlisting[linerange=160-170]{code/src/papercode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR
\caption{Reuse two language components \lstinline{VarExpr} and \lstinline{TypedLam}.}\label{fig:compose-components}
\end{figure}

### 2017-04-24 SLE_17/sections/Inheritance.tex
Revise code (up to section 4) to fit the narrower column

\lstinputlisting[linerange=29-31]{code/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_SIMPLE_LAM

@@

\lstinputlisting[linerange=35-38]{code/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_BAD_ATTEMPT

@@

\lstinputlisting[linerange=42-47]{code/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_APPROACH

Now \inlinecode{VarExprParser} successfully represents the parser for the extended language, because Scala uses dynamic dispatch for
method overriding in inheritance. When the input \inlinecode{"1 + x"} is fed to the parser \inlinecode{this.pExpr}, it firstly delegates
the work to \inlinecode{super.pExpr}, which parses literals and additions. However, the recursive call \inlinecode{pExpr} in \inlinecode{pAdd}
actually refers to \inlinecode{this.pExpr} again due to dynamic dispatch, and it covers the variable case. Similarly, all recursive calls can be updated to include new extensions if needed.

@@

\lstinputlisting[linerange=67-73]{code/Sec3Inheritance/Code1.scala}% APPLY:linerange=MULTIPLE_INHERITANCE

@@

\lstinputlisting[linerange=79-95]{code/Sec3Inheritance/Code1.scala}% APPLY:linerange=BOOL_EXAMPLE

>>>>

\lstinputlisting[linerange=7-9]{code/src/papercode/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_SIMPLE_LAM

@@

\lstinputlisting[linerange=13-16]{code/src/papercode/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_BAD_ATTEMPT

@@

\lstinputlisting[linerange=20-26]{code/src/papercode/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_APPROACH

Now \inlinecode{VarParser} successfully represents the parser for the extended language, because Scala uses dynamic dispatch for
method overriding in inheritance. When the input \inlinecode{"1 + x"} is fed to the parser \inlinecode{this.pExpr}, it firstly delegates
the work to \inlinecode{super.pExpr}, which parses literals and additions. However, the recursive call \inlinecode{pExpr} in \inlinecode{pAdd}
actually refers to \inlinecode{this.pExpr} again due to dynamic dispatch, and it covers the variable case. Similarly, all recursive calls can be updated to include new extensions if needed.

@@

\lstinputlisting[linerange=46-50]{code/src/papercode/Sec3Inheritance/Code1.scala}% APPLY:linerange=MULTIPLE_INHERITANCE

@@

\lstinputlisting[linerange=56-73]{code/src/papercode/Sec3Inheritance/Code1.scala}% APPLY:linerange=BOOL_EXAMPLE

### 2017-04-24 SLE_17/sections/ObjectAlgebras.tex
Revise code (up to section 4) to fit the narrower column

\lstinputlisting[linerange=23-23]{code/Sec4OA/Code1.scala}% APPLY:linerange=FVARSEXPR

\noindent Then all classes representing language constructs could be 
extended to implement the operation. A first well-known problem is that such
approach is problematic in terms of type-safety (but see recent work 
by Wang and Oliveira~\cite{wang2016expression}, which shows a technique that is type-safe
in many cases). More importantly, a second problem is that
\begin{comment}
But then programmers have to define a new class for each construct by
extending the old class with \lstinline{FVarsExpr}, and implementing
the corresponding \lstinline{fVars} method:

\lstinputlisting[linerange=27-37]{code/Sec4OA/Code1.scala}% APPLY:linerange=FVAR_CLASSES

Besides introducing boilerplate code, more classes have to be defined
along with future extensions. They pollutes the namespace heavily.
More importantly, 
\end{comment}
is that even if that approach would work, the parsing code would not
be reusable. \lstinline{VarExprParser} is no longer suitable for parsing the new ASTs! All the methods in the previous parser have type \inlinecode{Parser[Expr]}, which indicates the result has type \inlinecode{Expr}, whereas the type should be \inlinecode{FVarsExpr} now.

Therefore, in order to generate new AST structures, we have to
substitute \lstinline{FVarsExpr}, \lstinline{FVarsLit},
\lstinline{FVarsAdd} and \lstinline{FVarsVar} for the old
\lstinline{Expr}, \lstinline{Lit}, \lstinline{Add} and \lstinline{Var}
correspondingly. Thus we have to either modify existing code or 
use casts. For semantic modularity, we need a different approach for building ASTs.

%\bruno{The following text is not very good, needs to be revised. I think
%we need to show the full code and where it breaks. Since I removed text
%about the Expression problem in S3, I think we need to talk more about it %here.}
%It refers to one Scala solution to the Expression Problem.\huang{what's the %solution? does it have the later "critical issue"? if so, why is it a %"solution"} But then programmers have to define new classes for literals, %additions and variables
%by extending the old ones together with \inlinecode{FreeVarsExpr}. The %critical issue is that, the existing parsing code only generates objects of %the old
%types, again it is a modularity issue, since it is a bad idea to modify %existing code again and again with extensions. Other solutions including %modelling constructors as functions introduce unnecessary complexity.

@@

\lstinputlisting[linerange=5-8]{code/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALG
We have two constructors \inlinecode{lit} and \inlinecode{add}, corresponding to literals and additions. The trait \inlinecode{ExprAlg} is called an \textit{Object Algebra interface}. The trait is parametrized by the type
\inlinecode{E}, which abstracts over the concrete type of the AST.

@@

\lstinputlisting[linerange=12-15]{code/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_PRINT
Here \inlinecode{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
One can define evaluation operation as a new trait on this expression language:

\lstinputlisting[linerange=19-22]{code/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EVAL

\paragraph{Adding New AST Constructs}
Furthermore, new language constructs can be added by extending \inlinecode{ExprAlg} and adding new cases only. Now we extend the language
with variables. A new Object Algebra interface \inlinecode{VarAlg} is defined as follows:

\lstinputlisting[linerange=26-28]{code/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALGEXT
Now pretty-printing on the new language can be realized without modifying existing code:

\lstinputlisting[linerange=32-34]{code/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
An observation is that only the new case is implemented for pretty-printing, and the others have been inherited.
In other words, existing code was reused and was not modified!

@@

\lstinputlisting[linerange=41-41]{code/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_MAKEEXP

Note how the construction of the abstract syntax happens through
the use of factory methods, instead of constructors.
To pretty-print the expression,
we pass an instance of \inlinecode{VarExprPrint} to \inlinecode{makeExp}:

\begin{lstlisting}
makeExp(new VarExprPrint {})
\end{lstlisting}
which results in \inlinecode{"(1 + x)"} as expected.

@@

\begin{comment}
\inlinecode{ExprAlg} is defined for a small language using Object Algebras. Since Object Algebras represent ``objects'' implicitly as functions like \inlinecode{ExprAlg[E] =>} \inlinecode{E} for abstract \inlinecode{E}, we first try to build the corresponding parser by type \inlinecode{ExprAlg[E] =>} \inlinecode{Parser[E]}.
\end{comment}

\lstinputlisting[linerange=8-16]{code/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser works fine, but it is not extensible! For example, we have demonstrated in Section~\ref{subsec:overriding} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, trying to do a similar method overriding for \inlinecode{pExpr} would require a type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]}, where the type of the \emph{extended} Object Algebra interface appears in \emph{contravariant} position. This difference prohibits method/field overriding. Scala supports type-refinement of fields if the overriding
field has a type which is a \emph{subtype} of the original field.
Unfortunately, in this case, although \inlinecode{VarExprAlg} is a subtype
of \inlinecode{ExprAlg}, the function type \inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]} is a supertype of \inlinecode{ExprAlg[E] =>} \inlinecode{Parser[E]}. Thus it is not possible to have overriding.

% because Object Algebra interfaces extended from \inlinecode{VarExprAlg} %are subtypes of it, but the types of their parsers are supertypes of %\inlinecode{VarExprAlg[E] =>} \inlinecode{Parser[E]}.

\paragraph{A Solution}
A solution to this problem is instead having a field with the Object Algebra
in \lstinline{ExprOAParser}. This approach supports overriding as well.
The parsing code is as follows:

\lstinputlisting[linerange=23-30]{code/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER
This is precisely the pattern that we advocate for modular parsing.
One important remark is we introduce \inlinecode{pE} for recursive calls.
The reason why we use an extra and seemingly redundant field, is due to a subtle issue caused by Scala language and its parser combinator library. There is a restriction of \inlinecode{super} keyword in Scala that \inlinecode{super} can only use methods defined by keyword \inlinecode{def}, but cannot access fields defined by \inlinecode{val}, while the parser combinator library suggests using \inlinecode{val} to define parsers, especially for left-recursive ones. Our workaround is that we use different synonyms for \inlinecode{pE} in different traits, so that we can directly distinguish them by names without using \inlinecode{super}.

@@

\lstinputlisting[linerange=34-39]{code/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER

\noindent The type of the Object Algebra field \lstinline{alg} is first refined
to \lstinline{VarExpAlg[E]}, to allow calling the additional factory method
for variables. Note that, unlike the previous attempt, \lstinline{VarExpAlg[E]}
is a subtype of \lstinline{ExpAlg[E]}, thus the type-refinement of the
field is allowed. Now, the code for parsing variables (\lstinline{pVar}) can
call \lstinline{alg.varE}. The following code illustrates how to use
the parser from a client's perspective:

\lstinputlisting[linerange=43-46]{code/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER_CLIENT

In the client code shown above, we first need to choose a concrete
Object Algebra to initialize the \lstinline{alg} field in \lstinline{VarExprOAParser}. Here we pick the
pretty-printing algebra \lstinline{VarExprPrint}, but any other Object
Algebra that implements \lstinline{VarExprAlg} would work.
With an instance of \lstinline{VarExprOAParser} in hand, we can call
\lstinline{pE} to obtain the parser to feed to the \lstinline{parse} method.
The result shows that such a pattern provides modular parsing.

@@

\begin{comment}
\subsection{ASTs and Object Algebras}\label{subsec:more-oa}

In Section~\ref{subsec:objectalgebras}, \lstinline{makeExp} represents
the AST of expression \inlinecode{1 + x} as a method. Parsers with
Object Algebras generate such ``implicit objects'', which are actually
functions of type \lstinline{Alg[E] => } \lstinline{E} for generic
\lstinline{E}. Comparing with traditional objects, one may have worry
that if a parser produces function structures, it could be hard to
further process them. Especially, in the design of a compiler there
will be a lot of desugarings (or AST transformations). As previous
work on Object Alegbras has shown~\cite{Gouseti2014,Zhang2015}, AST
transformations can be implemented using \emph{transformation
  algebras}. Continuing with the previous example, suppose we want to
transform an expression of \inlinecode{VarExprAlg} to an expression of
\inlinecode{ExprAlg}, by replacing some variables with values based on
a variable environment. A \textit{transformation} algebra is
implemented as follows:

\lstinputlisting[linerange=45-56]{code/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_REFACTOR
Here the \inlinecode{VarExprAlg} expression \inlinecode{1 + x} is changed to \inlinecode{ExprAlg} expression \inlinecode{1 + 2}, before applying the pretty-printing algebra. The abstract field
\inlinecode{alg} in \inlinecode{Refactor} could be any algebra simply for delegation. With an algebra instance of \inlinecode{alg}, \inlinecode{Refactor} is able to transform a \inlinecode{VarExprAlg} expression before producing the result. Furthermore, transformations can be composed in a linear pipeline-style beforehand~\cite{Zhang2015}.

Another useful technique is the composition of \textit{query} algebras. A query traverses the abstract syntax tree, collects information and returns the result. The aforementioned \lstinline{Print} and \lstinline{Eval} algebras are traditional queries. But as we discussed before, an algebra is first fed to the parser, then \lstinline{parse} is invoked to do the parsing. It is tedious to fed \lstinline{Print} and \lstinline{Eval} separately, as the same input is parsed twice. Instead, we can firstly compose two queries into one~\cite{Oliveira:2012,oliveira2013feature}, before applying it for parsing.
\end{comment}

>>>>

\lstinputlisting[linerange=23-23]{code/src/papercode/Sec4OA/Code1.scala}% APPLY:linerange=FVARSEXPR

\noindent Then all classes representing language constructs could be
extended to implement the operation. A first well-known problem is that such
approach is problematic in terms of type-safety (but see recent work
by Wang and Oliveira~\cite{wang2016expression}, which shows a technique that is type-safe
in many cases). More importantly, a second problem is that
even if that approach would work, the parsing code would not
be reusable. \lstinline{VarParser} is no longer suitable for parsing the new ASTs! All the methods in the previous parser have type \inlinecode{Parser[Expr]}, which indicates the result has type \inlinecode{Expr}, whereas the type should be \inlinecode{NewExpr} now.

Therefore, in order to generate new AST structures, we have to
substitute \lstinline{NewExpr}, \lstinline{NewLit},
\lstinline{NewAdd} and \lstinline{NewVar} for the old
\lstinline{Expr}, \lstinline{Lit}, \lstinline{Add} and \lstinline{Var}
correspondingly. Thus we have to either modify existing code or
use casts. For semantic modularity, we need a different approach for building ASTs.

@@

\lstinputlisting[linerange=5-8]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALG
We have two constructors \inlinecode{lit} and \inlinecode{add}, corresponding to literals and additions. The trait \inlinecode{Alg} is called an \textit{Object Algebra interface}. The trait is parametrized by the type
\inlinecode{E}, which abstracts over the concrete type of the AST.

@@

\lstinputlisting[linerange=12-16]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_PRINT
Here \inlinecode{Print} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a string as the result.
One can define evaluation operation as a new trait on this expression language:

\lstinputlisting[linerange=20-23]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EVAL

\paragraph{Adding New AST Constructs}
Furthermore, new language constructs can be added by extending \inlinecode{Alg} and adding new cases only. Now we extend the language
with variables. A new Object Algebra interface \inlinecode{VarAlg} is defined as follows:

\lstinputlisting[linerange=27-29]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_ALGEXT
Now pretty-printing on the new language can be realized without modifying existing code:

\lstinputlisting[linerange=33-35]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_EXTPRINT
An observation is that only the new case is implemented for pretty-printing, and the others have been inherited.
In other words, existing code was reused and was not modified!

@@

\lstinputlisting[linerange=42-43]{code/src/papercode/Sec4OA/Code2.scala}% APPLY:linerange=OVERVIEW_OA_MAKEEXP

Note how the construction of the abstract syntax happens through
the use of factory methods, instead of constructors.
To pretty-print the expression,
we pass an instance of \inlinecode{VarPrint} to \inlinecode{makeExp}:

\begin{lstlisting}
makeExp(new VarPrint {})
\end{lstlisting}
which results in \inlinecode{"(1 + x)"} as expected.

@@

\lstinputlisting[linerange=8-17]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser looks fine, but it is not extensible. For example, we have demonstrated in Section~\ref{subsec:overriding} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, trying to do a similar method overriding for \inlinecode{pExpr} would require a type \inlinecode{VarAlg[E] =>} \inlinecode{Parser[E]}, where the type of the \emph{extended} Object Algebra interface appears in \emph{contravariant} position. This difference prohibits method/field overriding. Scala supports type-refinement of fields if the overriding
field has a type which is a \emph{subtype} of the original field.
Unfortunately, in this case, although \inlinecode{VarAlg} is a subtype
of \inlinecode{Alg}, the function type \inlinecode{VarAlg[E] =>} \inlinecode{Parser[E]} is a supertype of \inlinecode{Alg[E] =>} \inlinecode{Parser[E]}. Thus it is not possible to have overriding.

\paragraph{A Solution}
A solution to this problem is instead having a field with the Object Algebra
in \lstinline{OAParser}. This approach supports overriding as well.
The parsing code is as follows:

\lstinputlisting[linerange=21-30]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER
This is precisely the pattern that we advocate for modular parsing.
One important remark is we introduce \inlinecode{pE} for recursive calls.
The reason why we use an extra and seemingly redundant field, is due to a subtle issue caused by Scala language and its parser combinator library. There is a restriction of \inlinecode{super} keyword in Scala that \inlinecode{super} can only use methods defined by keyword \inlinecode{def}, but cannot access fields defined by \inlinecode{val}, while the parser combinator library suggests using \inlinecode{val} to define parsers, especially for left-recursive ones. Our workaround is that we use different synonyms for \inlinecode{pE} in different traits, so that we can directly distinguish them by names without using \inlinecode{super}.

@@

\lstinputlisting[linerange=34-39]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER

\noindent The type of the Object Algebra field \lstinline{alg} is first refined
to \lstinline{VarAlg[E]}, to allow calling the additional factory method
for variables. Note that, unlike the previous attempt, \lstinline{VarAlg[E]}
is a subtype of \lstinline{Alg[E]}, thus the type-refinement of the
field is allowed. Now, the code for parsing variables (\lstinline{pVar}) can
call \lstinline{alg.varE}. The following code illustrates how to use
the parser from a client's perspective:

\lstinputlisting[linerange=43-46]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER_CLIENT

In the client code shown above, we first need to choose a concrete
Object Algebra to initialize the \lstinline{alg} field in \lstinline{VarOAParser}. Here we pick the
pretty-printing algebra \lstinline{VarPrint}, but any other Object
Algebra that implements \lstinline{VarAlg} would work.
With an instance of \lstinline{VarOAParser} in hand, we can call
\lstinline{pE} to obtain the parser to feed to the \lstinline{parse} method.
The result shows that such a pattern provides modular parsing.

@@

### 2017-04-24 SLE_17/sections/PackratParsing.tex
Revise code (up to section 4) to fit the narrower column

\begin{figure}[t]
\centering
\lstinputlisting[linerange=42-53]{code/Sec2Packrat/Code2.scala}% APPLY:linerange=PACKRAT_PAPERCODE
\caption{Helper object for code demonstration in this paper.}\label{fig:papercode}
\end{figure}

@@

\lstinputlisting[linerange=9-17]{code/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_OO_AST

@@

\lstinputlisting[linerange=21-29]{code/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_SIMPLE_EXPR

In the trait \lstinline{ExprParser}, \lstinline{lexical} is used for lexing. \lstinline{pLit} parses an integer for the literal case.
\lstinline{pAdd} handles the addition case and creates an object of \lstinline{Add}. It parses two sub-expressions by calling \lstinline{pExpr}
recursively. Finally \lstinline{pExpr} composes \lstinline{pLit} and \lstinline{pAdd} using the longest match alternative combinator \inlinecode{|||}.
Table~\ref{tab:packrat} shows common parser combinators from the library.

@@

The last line demonstrates how we parse a valid expression \inlinecode{1 + 2} using our parser.

\lstinputlisting[linerange=33-33]{code/Sec2Packrat/Code2.scala}% APPLY:linerange=PACKRAT_RUNPARSER

>>>>

\begin{figure}[t]
\centering
\lstinputlisting[linerange=16-27]{code/src/papercode/Sec2Packrat/Code2.scala}% APPLY:linerange=PACKRAT_PAPERCODE
\caption{Helper object for code demonstration in this paper.}\label{fig:papercode}
\end{figure}

@@

\lstinputlisting[linerange=9-17]{code/src/papercode/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_OO_AST

@@

\lstinputlisting[linerange=21-28]{code/src/papercode/Sec2Packrat/Code1.scala}% APPLY:linerange=PACKRAT_SIMPLE_EXPR

In the trait \lstinline{AParser}, \lstinline{lexical} is used for lexing. \lstinline{pLit} parses an integer for the literal case.
\lstinline{pAdd} handles the addition case and creates an object of \lstinline{Add}. It parses two sub-expressions by calling \lstinline{pExpr}
recursively. Finally \lstinline{pExpr} composes \lstinline{pLit} and \lstinline{pAdd} using the longest match alternative combinator \inlinecode{|||}.
Table~\ref{tab:packrat} shows common parser combinators from the library.

@@

The code below demonstrates how to parse a valid expression \inlinecode{1 + 2} using our parser.

\lstinputlisting[linerange=6-7]{code/src/papercode/Sec2Packrat/Code2.scala}% APPLY:linerange=PACKRAT_RUNPARSER
